---
title: "Emotional Contagion Substantive Analyses"
author: "Nhung Vu"
date: "4/14/2022"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

# Brief Literature Review & Hypotheses

When faced with a problem, people commonly disclose personal information
and affect to receive social support. During supportive interactions,
the support recipient usually have to disclose their stressor in order
to receive support. Disclosure of stressors often include disclosure of
negative affect. Exposure to negative affect can drain support providers
because they may catch the distressed person's negative affect. This
transmission of emotions is often referred to as emotional contagion.
Emotional contagion occurs when people converge emotionally (Hatfield et
al., 1992).

Emotional contagion is especially important in supportive communication
because providers' emotions can influence how much and the quality of
support they provide. Emotions have a motivational component, called
action tendencies, which prompt people to engage in certain behavioral
responses (Frijda et al., 1989; Lazarus, 1991). In other words, a
person's affective state influences how they communicate and
experiencing emotional contagion can also influence people's
communicative behaviors. In the context of supportive communication, a
lack of provider resources (i.e., cognitive or emotional resources) may
limit support provision (Feeney & Collins, 2001, 2003). Thus, it is
possible that if listening to recipient's disclosure of stressors makes
providers feel worse, they will be less motivated to provide support.

H1: Changes in providers' positive affect and negative affect predict
their motivation to provide support.

Previous research suggested that emotions can be transmitted through
nonverbal behaviors. However, the presence of emotional contagion via
mediated communication channels, where access to nonverbal cues can be
limited depending on the channel, has also been documented (for a
review, see Goldernberg & Gross, 2020). Media richness theory (Daft and
Lengel, 1986) claimed that different communication channels varied in
their ability to reproduce "rich" social information (e.g., immediate
feedback, language variety, personalization, number of cues), thereby
making some media more appropriate than others for certain tasks. In
other words, the richness of a communication channel is determined by
its ability to offer rapid feedback, multiple cues, natural language,
and personal focus. Under this view, communication channels can be
arranged on a continuum from lean to rich. Specifically, video chat
would be considered a richer medium than text and audio chat because
compared to text and audio chat, people can provide rapid feedback and
transmit multiple cues more easily through video chat. Furthermore,
emotional contagion is more likely to occur in richer channels, such as
video chat, because a larger amount of social information, including
emotional cues, can be transmitted through these channels. Thus, this
study proposes:

H2: Participants in the video chat condition will experience a greater
increase in negative affect and a greater decrease in positive affect
after the interaction compared to those in the audio chat and text
conditions.

H3: Participants in the audio chat condition will experience a greater
increase in negative affect and a greater decrease in positive affect
after the interaction compared to those in the text condition.

# Description of Data Set

## Study design

The current data set is collected from an experiment conducted over
Zoom. Participants were asked to bring a conversational partner to the
lab. Upon arrival, participants were be randomly assigned to either the
text condition, the voice call condition, or the video chat condition.
Within each dyad, individuals were be randomly assigned so that a person
will be the support provider and the other one will be the support
recipient. Participants were instructed to fill out a pre-interaction
survey, interact with their conversational for 15 minutes, then fill out
a post-interaction survey.

## Description of data set

The data set has two components. Participants' responses to the
pre-interaction survey are recorded in the "ECPre.csv" file.
Participants' responses to the post-interaction survey are recorded in
the "ECPost.csv" file.

In both surveys, participants were asked to provide the unique ID they
were given at the outset of the experiment (i.e., when they first joined
the researcher's Zoom room). Participants were also asked to indicate
the role they were assigned to. In the data set, participants who were
assigned to be providers received a "1" and participants who were
assigned to be recipients received a "2" for the variable "role".

The ECPre and the ECPost files also contain the following measures that
are relevant to the current analysis (other variables in the data files
that are not used in this project will not be reported here):

Participants affect in the moment were be measured with the 20-item
Positive and Negative Affect Schedule (PANAS; Watson, Clark, & Tellegen,
1988). Participants were asked to indicate how much they experience a
variety of the emotional states (e.g., joy, love, sadness) on a
five-point Likert type scale (1 = very slightly or not at all, 5 =
extremely). The affect items in the pre-interaction survey are labeled
**panasPre** and the affect items in the post-interaction survey are
labeled **panasPost**. Items 1, 3, 5, 9, 10, 12, 14, 16, 17, and 19 were
used to measure positive affect. Items 2, 4, 6, 7, 8, 11, 13, 15, 18,
and 20 were used to measure negative affect. (Previous research sum the
items instead of averaging the items to create composite variables, so I
will sum the items for the composite variables in this project as well).

Support providers' motivation to provide support was measured after the
interaction with a 5-item scale adapted from High and Scharp (2015).
Participants responded to items on a 5-point Likert scale (1 = strongly
disagree, 5 = strongly agree). These scale items were displayed only to
participants who indicated in the survey that they were support
providers during the interaction portion. The items of this scale are
labeled **motivationP** in the data set.

Finally, the last column of the ECPost data file contains information
regarding which experimental conditions (i.e., video, voice, or text)
participants were assigned to.

Next, I am going to create composite variables and calculate variables
needed to test the hypotheses above. Specifically, the label for the
composite variables will be:

**paPre** is positive affect pre interaction **nePre** is negative
affect pre interaction **paPost** is positive affect post interaction
**naPost** is negative affect post interaction **paDiff** is the
difference in positive affect between Pre and Post interaction (paPost -
paPre) **naDiff** is the difference in negative affect between Pre and
Post interaction (naPost - naPre) **motivationP** is provider's
motivation to provide support (motivationP)

```{r, warning = FALSE}
library(psych)
library(dplyr)

predata=read.csv("ECPre.csv", na.strings = NA) #read in the pre-interaction data file
postdata=read.csv("ECPost.csv", na.strings = NA) #read in the post-interaction data file
ecdata=merge(predata,postdata,by=c("ID","role")) #merge the two data files

#create composite variables
ecdata <- ecdata %>%
  mutate(paPre = (panasPre_1+panasPre_3+panasPre_5+panasPre_9+panasPre_10+panasPre_12+panasPre_14+panasPre_16+panasPre_17+panasPre_19),
        naPre = (panasPre_2+panasPre_4+panasPre_6+panasPre_7+panasPre_8+panasPre_11+panasPre_13+panasPre_15+panasPre_18+panasPre_20),
        paPost = (panasPost_1+panasPost_3+panasPost_5+panasPost_9+panasPost_10+panasPost_12+panasPost_14+panasPost_16+panasPost_17+panasPost_19),
        naPost = (panasPost_2+panasPost_4+panasPost_6+panasPost_7+panasPost_8+panasPost_11+panasPost_13+panasPost_15+panasPost_18+panasPost_20),
        paDiff=(paPost-paPre),
        naDiff=(naPost-naPre),
        motivationP=(motivationP_1+motivationP_2+motivationP_3+motivationP_4+motivationP_5)/5)

#get descriptive statistics for the variables
describe(ecdata$paPre)
describe(ecdata$naPre)
describe(ecdata$paPost)
describe(ecdata$naPost)
describe(ecdata$paDiff)
describe(ecdata$naDiff)
describe(ecdata$motivationP)
```

## Check for abnormalities

```{r, warning=FALSE}
library(GGally)
ggpairs(ecdata[,224:231])
```

The variable **motivationP** is negatively skewed, which could be a
problem. Besides that, there is nothing that is obviously unusual in the
other variables of interest based on the ggpairs plot. The variables
**paDiff** and **naDiff** are generally normally distributed.

To resolve the skewness in the variable "motivationP," I will log
transform the variable

```{r}
#histogram for raw motivationP variable
hist(filter(ecdata,role==1)$motivationP)

#log transform motivationP variable
ecdata$motivationP_log = log(ecdata$motivationP)
hist(filter(ecdata,role==1)$motivationP_log)
```

The log of **motivationP** still does not look normally distributed. So,
I will run two regression equations to test the first hypothesis where
motivation is a variable of interest. The fist regression will use the
untransformed motivation variable (i.e., "motivationP") and the second
regression will use the log of motivation variable (i.e.,
"motivationP_log").

# Analyses

## H1: Changes in providers' positive affect and negative affect predict their motivation to provide support.

To test this hypothesis, I am going to construct a multiple regression
equation with providers' change in positive and negative affect as
predictors. I am choosing to multiple regression here because I want to
estimate the relationship between two predictors and a dependent
variable. I could use two simple linear regression equations to estimate
the effect of positive affect on motivation and negative affect on
motivation separately; however, based on the literature on affect, the
assumption that positive affect and negative affect independently
influence an outcome is unrealistic. Therefore, I entered both
predictors into one equation.

Furthermore, providers' positive and negative affect are also included
in the equation as control variable because providers' affect before
interaction could influence their motivation to seek support. To ensure
the influence of these variables on the outcome is accounted for, they
are included in the model.

Finally, positive affect and negative affect will not be allowed to
interact because previous research has suggested that positive affect
and negative affect are orthogonal dimensions of affect (Watson, 1988).
In other words, no interaction term will be entered into the equation.

```{r, warning = FALSE}
#create models to test H1

#raw motivation as outcome
H1_model=lm(motivationP~paPre+naPre+paDiff+naDiff,data=filter(ecdata,role==1)) 
summary(H1_model)

#transformed motivation as outcome
H1tf_model=lm(motivationP_log~paPre+naPre+paDiff+naDiff,data=filter(ecdata,role==1)) 
summary(H1tf_model) 

#create diagnostic plots to check for violations of MLR assumptions
#model with raw motivation
plot(H1_model)
#model with transformed motivation
plot(H1tf_model)

confint(H1_model,level=0.95)
```

Note: The diagnostic plots for the model in which the raw motivation
variable (i.e., motivationP) was used as outcome are very similar to the
diagnostic plots for the model in which the log transformed motivation
variable (i.e., motivationP_log) was used as the outcome variable. In
other words, the distribution of the residuals is similar for both
models. For ease of interpretation, I am going to move forward with the
untransformed model.

Comments on diagnostic plots:

Residuals vs. Fitted plot: The red line is roughly horizontal. There's
some curvature to the left of the plot, but given the small sample size,
this is fine for now.

Normal QQ plot: There is noticeable deviation from the diagonal line
between the first and the second theoretical quantiles. Thus, residuals
may not be normally distributed.

Scale-Location plot: Generally, this plot looks normal. The red line is
roughly horizontal. Point #36 may be an outlier.

Residuals vs. Leverage plot: There may be some high leverage points in
the data set. However, they still fall within acceptable Cook's distance
range, so I am going to retain them.

Overall, there is no obvious violation of multiple linear regression
assumptions. So, I am going to move forward with the results above.

## H2 & H3

**H2**: Participants in the video chat condition will experience a
greater increase in negative affect and a greater decrease in positive
affect after the interaction compared to those in the audio chat and
text conditions.

**H3**: Participants in the audio chat condition will experience a
greater increase in negative affect and a greater decrease in positive
affect after the interaction compared to those in the text condition.

To test H2 and H3, first, I will dummy code the **channel** variable. I
choose the dummy coding scheme here because I want to compare conditions
with one another. Effect coding will let me compare groups' means with
the grand mean, which is not a comparison of interest in the current
project. I could also use contrast coding here; however, H2 and H3 make
very similar comparisons and dummy coding can answer both questions
without introducing complexity into the coding scheme.

After dummy coding the variable **channel**, I am going to run 2
multiple regression models to test H2 and H3. The first model will
estimate the effect of channel on positive affect, and the second model
will estimate the effect of channel on negative affect. I am running
multiple regression here because there are more than one predictor in
each model. Again, participants' affect before the interaction will be
entered into the equations as a control variable. Then, two dummy
variables (i.e., isVideo and isVoice) for channel will be entered as
predictors.

Because these first two models will not provide the comparison between
the video and voice conditions, I am going to make the voice condition
the reference group in the next two models.

```{r, warning = FALSE}
#dummy coding channel variable
ecdata$isVideo = as.integer(ecdata$channel == "video")
ecdata$isVoice = as.integer(ecdata$channel == "voice")
ecdata$isText = as.integer(ecdata$channel == "text")

#model with positive affect as outcome (comparisons tested: video vs. text and voice vs. text)
H2pa_model=lm(paDiff~paPre+isVideo+isVoice,data=ecdata)
summary(H2pa_model)
confint(H2pa_model,level=0.95)

#model with negative affect as outcome (comparisons tested: video vs. text and voice vs. text)
summary(lm(naDiff~naPre+isVideo+isVoice,data=ecdata))
confint(lm(naDiff~naPre+isVideo+isVoice,data=ecdata),level=0.95)

#model with positive affect as outcome(comparison tested: video vs. voice)
summary(lm(paDiff~paPre+isVideo+isText,data=ecdata))
confint(lm(paDiff~paPre+isVideo+isText,data=ecdata),level=0.95)


#model with negative affect as outcome (comparison tested: video vs. voice)
summary(lm(naDiff~naPre+isVideo+isText,data=ecdata))
confint(lm(naDiff~naPre+isVideo+isText,data=ecdata),level=0.95)


```

# Results

The first hypothesis states that changes in providers' positive affect
and negative affect predict their motivation to provide support. To test
this hypothesis, a multiple linear regression was carried out.
Providers' motivation to provide support was entered as the dependent
variable. Provider's positive affect and negative affect before the
interaction were entered as control variables and the change in positive
affect and change in negative affect were entered as predictors. The
results of the regression indicated that the model explained 7.27% of
the variance and that the model was not significant, F(2,49) = 0.90,p =
0.4713. Providers' motivation to provide support was not predicted by
the changes their positive affect (β = 0.024, p = 0.103) or negative
affect (β = 0.019, p = 0.528). Thus, the first hypothesis was not
supported.

The second hypothesis states that participants in the video chat
condition will experience a greater increase in negative affect and a
greater decrease in positive affect after the interaction compared to
those in the audio chat and text conditions. Additionally, the third
hypothesis states that participants in the audio chat condition will
experience a greater increase in negative affect and a greater decrease
in positive affect after the interaction compared to those in the text
condition.To test these hypotheses, four multiple linear regression was
conducted.

In the first regression constructed to test H2 and H3, participants'
change in positive affect was entered as the outcome variable.
Participants' positive affect before the interaction was entered as a
control variable. Dummy variables created for the video and voice
conditions were entered as predictors. The results of the regression
indicated that the model explained 5.40% of the variance and that the
model was not significant, F(3,100) = 1.902,p = 0.1324. Whether
participants were in the text, voice, or video condition did not predict
the change in their positive affect. The difference in the change in
positive affect between the voice and the text condition was not
significant (β = 0.49, p = 0.72). The difference in the change in
positive affect between the video and the text condition was also not
significant (β = 0.82, p = 0.54).

In the second regression constructed to test H2 and H3, participants'
change in negative affect was entered as the outcome variable.
Participants' negative affect before the interaction was entered as a
control variable. Dummy variables created for the video and voice
conditions were entered as predictors. The results of the regression
indicated that the model explained 44.5% of the variance and that the
model was significant, F(3,99) = 26.44,p \< 0.001. However, whether
participants were in the text, voice, or video condition did not predict
the change in their negative affect. The difference in the change in
negative affect between the voice and the text condition was not
significant (β = 0.63, p = 0.42). The difference in the change in
negative affect between the video and the text condition was also not
significant (β = 0.08, p = 0.92).

The third and fourth regression equations that were constructed to test
H2 and H3 focus on testing the comparisons between the video condition
against the voice condition. The difference in the change in negative
affect between the video and the voice condition was not significant (β
= 0.33, p = 0.80). The difference in the change in positive affect
between the video and the voice condition was also not significant (β =
0.54, p = 0.47). Overall, the second and the third hypotheses were not
supported.

# Interpretations of Results

The results suggest that changes in providers' positive and negative
affect did not predict their motivation to provide support. Furthermore,
the experimental conditions participants were assigned to did not
predict changes in their positive or negative affect from before the
interaction to after the interaction. It is possible that other factors
influence participants' changes in affect. It is also possible that the
current sample size is not large enough to detect an effect. (note: this
is an ongoing project, we are still collecting data)

## Describe what results would have meant if they were significant.

The results from the first regression model (i.e., model constructed to
test H1) would suggest that changes in providers' positive and negative
affect predict their motivation to provide support. Specifically,
providers' positive affect would be positively associated with their
motivation to provide support while holding the other variables
constant. Similarly, providers' negative affect would be positively
associated (based on the coefficient calculated above) with their
motivation to provide support while holding the other variables
constant.

The results from models testing H2 and H3 would suggest that
participants' changes in positive affect and negative affect were
predicted by the experimental conditions they were assigned to.
Specifically, participants in the video condition would experience a
greater increase in both positive and negative affect after the
interaction compared to participants in the text condition. Participants
in the voice condition would also experience a greater increase in both
positive and negative affect after the interaction compared to
participants in the text condition.

For the models that were constructed so that the voice condition was the
reference group, we would see that participants in the video condition
would experience a greater increase in both positive and negative affect
after the interaction compared to participants in the voice condition.

# Plots

## H1 results

```{r, warning = FALSE}
library(interactions)
interact_plot(H1_model, paDiff, naDiff, modxvals = NULL, interval = TRUE, int.type = "confidence", int.width = 0.95, outcome.scale = "response", x.label = "change in positive affect", y.label = "motivation to provide support", modx.labels = "change in negative affect", main.title = "Hypothesis 1", legend.main = "change in negative affect", color.class = NULL, line.thickness = 1.1, vary.lty = TRUE)
```

## H2 & H3 results

```{r, warning = FALSE}
library(ggplot2)
ecdata$channel=as.factor(ecdata$channel)
ecdata %>%
 ggplot(aes(x = paPre, y = paDiff, color = channel))+
 geom_point()+
 geom_smooth(method = "lm")+ #channel-specific regression lines
  xlab("positive affect before interaction")+ylab("change in positive affect")+
  ggtitle("Hypotheses 2 and 3 (positive affect)")
```

```{r, warning=FALSE}
ecdata %>%
 ggplot(aes(x = naPre, y = naDiff, color = channel))+
 geom_point()+
 geom_smooth(method = "lm")+ #channel-specific regression lines
  xlab("positive affect before interaction")+ylab("change in positive affect")+
  ggtitle("Hypotheses 2 and 3 (negative affect)")
```

# References

Daft, R. L., Lengel, R. H., & Trevino, L. K. (1987). Message
equivocality, media selection, and manager performance: Implications for
information systems. MIS Quarterly, 11, 355--366. <doi:10.2307/248682>

Feeney, B. C., & Collins, N. L. (2001). Predictors of caregiving in
adult intimate relationships: an attachment theoretical perspective.
Journal of personality and social psychology, 80(6), 972.

Feeney, B. C., & Collins, N. L. (2003). Motivations for caregiving in
adult intimate relationships: Influences on caregiving behavior and
relationship functioning. Personality and Social Psychology Bulletin,
29(8), 950-968.

Frijda, N. H., Kuipers, P., & ter Schure, E. (1989). Relations among
emotion, appraisal, and emotional action readiness. Journal of
Personality and Social Psychology, 57(2), 212--228.
<https://doi.org/10.1037/0022-3514.57.2.212>

Goldenberg, A., & Gross, J. J. (2020). Digital Emotion Contagion. Trends
in Cognitive Sciences, 24(4), 316--328.
<https://doi.org/10.1016/j.tics.2020.01.009>

Hatfield, E., Cacioppo, J. T., & Rapson, R. L. (1992). Primitive
emotional contagion. In M. S. Clark (Ed.), Emotion and social behavior
(pp. 151--177). Sage Publications, Inc.

High, A. C., & Scharp, K. M. (2015). Examining family communication
patterns and seeking social support direct and indirect effects through
ability and motivation. Human Communication Research, 41(4), 459-479.
<doi:10.1111/hcre.12061>

Lazarus, R. S. (1991). Cognition and motivation in emotion. American
Psychologist, 46, 352--367.

Watson, D. (1988). The vicissitudes of mood measurement: effects of
varying descriptors, time frames, and response formats on measures of
positive and negative affect. Journal of Personality and social
Psychology, 55(1), 128.

Watson, D., Clark, L. A., & Tellegen, A. (1988). Development and
validation of brief measures of positive and negative affect: The PANAS
scales. Journal of Personality and Social Psychology, 54(6), 1063--1070.
<https://doi.org/10.1037/0022-3514.54.6.1063>
